---
title: "Polygon-to-H3 compaction workflow"
description: "A reproducible workflow for polygon coverage, H3 compaction, source relationships, attribute aggregation, quality assurance, and cartographic representation."
vignette: >
  %\VignetteIndexEntry{Polygon-to-H3 compaction workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`h3compactR` provides a small workflow for representing polygon geography with
H3 cells, compacting those cells using established H3 hierarchy operations,
retaining the relationship between source and compacted cells, aggregating
attributes explicitly, and validating the result.

The central analytical rule is:

> The source H3 grid remains the authoritative analytical support. Compaction
> changes its representation; it does not change source-cell membership or
> invent analytical values.

This vignette uses one authoritative source grid throughout the complete
workflow.

```{r packages}
library(h3compactR)
library(sf)

data(toy_membership_polygons)
```

## 1. Prepare example polygons

The package example geography contains several polygon forms useful for
demonstrating assignment and compaction. For this vignette we create generic
feature identifiers and attributes directly from the example rows.

```{r prepare-polygons}
polygons <- toy_membership_polygons

polygons$feature_id <- sprintf(
  "feature_%02d",
  seq_len(nrow(polygons))
)

polygons$feature_name <- paste(
  "Feature",
  seq_len(nrow(polygons))
)

polygons$group_name <- rep(
  c("Group A", "Group B"),
  length.out = nrow(polygons)
)

polygons
```

The identifier used for assignment must uniquely identify polygon features.
The package validates polygon topology before performing polygon relationship
or assignment operations.

## 2. Create the authoritative source H3 grid

First create a single polygon footprint from the example features. The H3
coverage generated from this footprint is the authoritative source support
used for assignment, compaction, lookup, aggregation, and QA below.

```{r source-coverage}
polygon_union <- sf::st_union(
  sf::st_geometry(polygons)
)

polygon_union <- sf::st_sf(
  geometry = polygon_union
)

source_h3 <- h3_cover_polygon(
  x = polygon_union,
  resolution = 8L,
  boundary = "intersects"
)

length(source_h3)
```

### Coverage rules

`h3_cover_polygon()` supports four explicit boundary rules:

- `"center"` retains cells whose H3 centre is covered by the polygon;
- `"within"` retains cells fully contained by the polygon;
- `"intersects"` retains cells having any geometric intersection with the
  polygon; and
- `"overlap"` retains cells meeting a specified proportional overlap
  threshold.

For overlap coverage, the retained proportion is

\[
\frac{\operatorname{area}(\mathrm{H3\ cell}\cap\mathrm{polygon})}
     {\operatorname{area}(\mathrm{H3\ cell})}.
\]

For example:

```{r overlap-example, eval=FALSE}
overlap_h3 <- h3_cover_polygon(
  x = polygon_union,
  resolution = 8L,
  boundary = "overlap",
  min_overlap = 0.5
)
```

The choice of boundary rule is an analytical decision. Geometry-based rules
such as `within`, `intersects`, and `overlap` require more spatial processing
than standard centre-based coverage.

## 3. Optional neighbourhood expansion

`expand_h3()` adds neighbouring cells at the same H3 resolution. Expansion is
separate from polygon coverage so that the two operations remain explicit.

```{r expand}
expanded_h3 <- expand_h3(
  source_h3,
  rings = 1L
)

c(
  source = length(source_h3),
  expanded = length(expanded_h3)
)
```

The expanded grid is not used in the remaining workflow. The original
`source_h3` object remains authoritative.

## 4. Assign source cells to polygon features

Assignment is performed on the same source H3 cells created above.

```{r assignment}
source_assignment <- h3_assign_polygon(
  x = source_h3,
  polygons = polygons,
  id = "feature_id",
  keep = c("feature_name", "group_name"),
  outside = "unassigned"
)

head(source_assignment)

table(
  source_assignment$assignment_method,
  useNA = "ifany"
)
```

Assignment methods describe how ownership was established. Depending on the
geometry, these may include centre-based assignment, a unique intersection,
largest overlap, nearest assignment when requested, ambiguity, or an
unassigned result.

For this example the authoritative footprint should resolve to one polygon
owner per source H3 cell.

```{r assignment-check}
stopifnot(
  nrow(source_assignment) == length(source_h3),
  identical(sort(source_assignment$h3), sort(source_h3)),
  !any(source_assignment$assignment_method %in% c("ambiguous", "unassigned")),
  !anyNA(source_assignment$polygon_id)
)
```

## 5. Compact the source H3 representation

Compaction operates on the authoritative source indexes. `min_resolution`
defines the coarsest H3 resolution permitted in the compact representation.

```{r compact}
compacted_h3 <- compact_h3(
  x = source_h3,
  min_resolution = 7L
)

c(
  source = length(source_h3),
  compacted = length(compacted_h3)
)
```

The result may contain more than one H3 resolution. This is expected:
compaction replaces complete groups of child cells with their H3 parent where
the hierarchy permits it.

Compaction should therefore be understood as a hierarchical representation of
the source grid rather than a new analytical support.

## 6. Build the source-to-compact lookup

`h3_compaction_lookup()` records exactly which compact H3 cell owns each source
H3 cell.

```{r lookup}
lookup <- h3_compaction_lookup(
  source = source_h3,
  compacted = compacted_h3
)

head(lookup)

table(lookup$relationship)
```

Each authoritative source cell must occur exactly once in this relationship.

```{r lookup-check}
stopifnot(
  nrow(lookup) == length(source_h3),
  identical(sort(lookup$source_h3), sort(source_h3))
)
```

This lookup is the analytical bridge between the source grid and the compacted
representation.

## 7. Aggregate attributes safely

Additive measures can be summed through the lookup. Categorical attributes can
be collapsed to retain provenance.

Here we create a small deterministic additive measure on the source cells.

```{r source-data}
source_data <- source_assignment

source_data$observations <- rep(
  c(0, 1, 2, 1),
  length.out = nrow(source_data)
)

source_data$source_weight <- seq_len(
  nrow(source_data)
)

head(source_data)
```

Now aggregate using the lookup created from exactly the same authoritative H3
support.

```{r aggregation}
compact_data <- aggregate_h3(
  data = source_data,
  lookup = lookup,
  sum = c("observations", "source_weight"),
  collapse = c(
    "polygon_id",
    "feature_name",
    "group_name"
  ),
  h3_col = "h3"
)

head(compact_data)
```

`aggregate_h3()` does not infer aggregation semantics. Additive and categorical
fields are declared explicitly.

Collapsed categorical values record source provenance. They are not new
analytical categories. Where more than one unique source value contributes to
a compact owner, the values are retained in deterministic collapsed form and
the corresponding `*_n` field records the number of unique non-missing source
values represented.

### Check additive conservation

```{r conservation}
stopifnot(
  sum(source_data$observations) ==
    sum(compact_data$observations),
  sum(source_data$source_weight) ==
    sum(compact_data$source_weight)
)

c(
  source_observations = sum(source_data$observations),
  compact_observations = sum(compact_data$observations),
  source_weight = sum(source_data$source_weight),
  compact_weight = sum(compact_data$source_weight)
)
```

This conservation is fundamental: every authoritative source H3 contributes
its additive value exactly once to exactly one compact owner.

Derived quantities such as rates, proportions, and percentages should not
normally be averaged during compaction. Their appropriate numerator and
denominator components should be aggregated and the derived quantity
recalculated afterwards.

## 8. Validate the compaction

`qa_h3_compaction()` checks the structural relationship between the source and
compacted H3 representations.

```{r qa}
qa <- qa_h3_compaction(
  source = source_h3,
  compacted = compacted_h3
)

qa
```

The QA process checks round-trip reconstruction, hierarchy consistency,
source ownership, resolution composition, and the reduction achieved by
compaction.

## 9. Cartographic inspection

`plot_h3()` provides a lightweight way to inspect H3 geometry against polygon
context.

```{r plot-source}
plot_h3(
  source_h3,
  context = polygons,
  main = "Source H3 representation"
)
```

```{r plot-compact}
plot_h3(
  compacted_h3,
  context = polygons,
  main = "Compacted H3 representation"
)
```

Visual inspection is useful because a compact H3 parent may extend
geographically beyond the exact union of its represented source cells. H3
parent-child hierarchy is logically exact, but polygon geometry across
resolutions should not be interpreted as exact geographic containment.

For that reason, source membership and analytical values remain attached to
the authoritative source support and are transferred through the explicit
lookup rather than inferred from the visible footprint of a compact parent.

## 10. Polygon membership as a separate relationship

`h3_polygon_membership()` can be used when the required result is the
geometric relationship between H3 cells and polygon features rather than a
single assigned owner.

```{r membership}
membership <- h3_polygon_membership(
  x = source_h3,
  polygons = polygons,
  id = "feature_id"
)

head(membership)
```

Membership and assignment answer different questions:

- membership records H3-to-polygon relationships;
- assignment resolves one polygon owner for each H3 cell.

Keeping these operations separate avoids silently converting a geometric
relationship into an ownership rule.

## 11. Computational scale

Standard centre-based H3 coverage is substantially less computationally
demanding than geometry-based coverage rules. `within`, `intersects`, and
`overlap` require additional H3 geometry construction and spatial operations
and can become computationally expensive for very large areas at fine H3
resolutions.

There is no universal maximum cell count that is appropriate for every
workflow. Resolution and boundary rule should be selected according to the
analytical purpose, and large workflows should be tested on representative
subsets before scaling.

Performance optimisation for very large geometry-heavy workflows is an area
for future package development.

## 12. Analytical principles

The workflow demonstrated here follows several deliberate principles:

1. **The source H3 grid is authoritative.** Compaction changes representation,
   not source support.
2. **Compaction follows the H3 hierarchy.** It does not create a new H3
   compaction algorithm.
3. **Source-to-compact relationships are explicit.** Analytical values are not
   inferred from compact polygon appearance.
4. **Aggregation follows variable meaning.** Additive measures are summed;
   categorical fields may retain provenance; derived measures should be
   recalculated from appropriate components.
5. **QA is part of the workflow.** Round-trip reconstruction, ownership,
   hierarchy, conservation, and cartographic inspection provide complementary
   checks.
6. **Spatial relationships and ownership are separate concepts.** Membership
   does not silently imply assignment.

## Attribution

`h3compactR` builds on the H3 hierarchical geospatial indexing system and the
H3 operations exposed to R through `h3jsr`. Spatial geometry operations use
the `sf` ecosystem.

The package does not introduce a new H3 compaction algorithm. Its contribution
is the reproducible analytical workflow around polygon coverage choices,
hierarchical compaction, source-to-compact relationships, explicit attribute
aggregation, provenance, quality assurance, and cartographic representation.

The package's emphasis on making the choice and interpretation of geographic
units explicit is also informed by the broader cartographic literature on
geographic units and choropleth mapping, including Boscoe and Pickle (2003).

See the package references for full attribution and source details.
